Unsupervised Spectral Classification of DESI White Dwarfs by Dimensionality Reduction

### 2024-09-04 src/tex/manuscript.tex
5th draft

% Abstract of the paper
\begin{abstract}
As a new generation of large-sky spectroscopic surveys comes online, the enormous data volume poses unprecedented challenges in classifying spectra.
Although supervised classifiers exist, their complete reliance on training data can lead to overconfident predictions and bias against rarer classes.
We outline the use of dimensionality reduction to generate a 2D map of the structure of an intermediate-resolution spectroscopic dataset.
We demonstrate that this fully unsupervised technique efficiently separates white dwarfs of different spectral classes in the Dark Energy Spectroscopic Instrument's Early Data Release (DESI EDR), identifying spectral features that had been missed even by visual classification.
Requiring mere seconds of computation time, dimensionality reduction provides a useful automated complement to a human classifier.
By zooming in to important wavelength regions, we show how focusing the method on particular spectral features enables the identification of helium lines with 90 per cent accuracy, and cataclysmic variables with 100 per cent accuracy.
We also demonstrate a way of using dimensionality reduction in a supervised manner, outlining a procedure to classify any white dwarf spectrum in comparison with those in the DESI EDR.
With upcoming surveys promising tens of millions of spectra, this work highlights the potential for unsupervised techniques as efficient means of classification and dataset visualisation.
\end{abstract}

@@

A suite of upcoming Stage IV large-sky spectroscopic surveys promises to enormously increase the number of sources with intermediate-resolution spectroscopy, enabling advances across all areas of astronomy.
These surveys include:
\begin{itemize}
\item
\textbf{4MOST}
(4-metre Multi-Object Spectroscopic Telescope; \citealt{4most}): nine surveys which aim to obtain accurate radial velocities and chemical abundances for millions of Galactic stars, probe the history of supermassive black hole accretion through a spectroscopic sample of about 1 million active galactic nuclei, and constrain cosmological parameters by obtaining redshifts for 10-20 million emission line galaxies;
\item
\textbf{DESI}
(Dark Energy Spectroscopic Instrument; \citealt{desii, desiii}), which plans to study the growth of structure in the Universe through observations of baryon acoustic oscillations and redshift-space distortions, and trace the dark matter distribution by obtaining redshifts for 30 million galaxies and quasars at $1.0\lesssim z \lesssim 3.5$.
When these faint targets are obscured by moonlight over Iolkam Du'ag / Kitt Peak, the targeting switches to surveys of Milky Way stars and bright galaxies;
\item
\textbf{SDSS-V}
(Sloan Digital Sky Survey V; \citealt{sdssv}):
an all-sky survey from two $2.5\,\text{m}$ telescopes (one in each hemisphere), this aims to record multi-epoch spectra for over six million targets, including five million Galactic stars;
\item
\textbf{WEAVE}
(William Herschel Telescope Enhanced Area Velocity Explorer; \citealt{weave}): a northern-hemisphere complement to 4MOST, WEAVE will record accurate spectroscopic velocities of Galactic disk and halo stars, IFU H~\textsc{i} maps of $\sim10^4$ low-redshift galaxies, and wide-area views of galaxy clusters.
\end{itemize}
All four of these surveys have achieved first light, with SDSS-V's first data release (SDSS DR19; \citealt{sdssdr19}) and DESI's early data release (DESI EDR; \citealt{desiedr}) delivered last year.
As these spectroscopic surveys begin to return enormous quantities of data, and with the exabyte-scale Legacy Survey of Space and Time (LSST; \citealt{lsst}) on the horizon, automated techniques will become absolutely necessary in aiding the extraction of scientific results from the vast wealth of data being collected.

Entries in astronomical datasets often have high dimensionality.
They may be images of hundreds of pixels, light curves with thousands of measurements, or spectra with thousands of fluxes.
Any \textit{structure} in a dataset -- e.g., clusters or sequences -- can be investigated quantitatively by representing each entry as a vector in a very high-dimensional space: a dataset of $N$ spectra with $D$ flux bins can be thought of as a set of $N$ points in $D$-dimensional data space.
Visualisation of datasets in more than two or three dimensions is challenging, but whereas a $D$-dimensional data point might ostensibly live in $\mathbb{R}^D$, it is very likely that every point in the dataset is located on or near a much smaller submanifold within this space.
In a spectrum, the flux in a given bin will usually be similar to the flux in adjacent bins.
Similarly, two different sources of the same astrophysical type (e.g.\ two white dwarfs of the same spectral classification) will have similar spectra, and will thus be nearby in the data space.
This massively reduces the effective dimensionality of the dataset, compared to the dimensionality of the raw data points themselves.
The observation that many datasets exist near low-dimensional submanifolds of the data space has motivated the development of several \textit{dimensionality reduction} (DR) techniques, including Principal Component Analysis (PCA; \citealt{pca}), diffusion maps \citep{coifman06, lafon06}, Locally Linear Embedding (LLE; \citealt{roweis00}), $t$-distributed Stochastic Neighbour Embedding ($t$SNE; \citealt{tsne}), and Uniform Manifold Approximation and Projection (UMAP; \citealt{umap}).
These methods calculate a two-dimensional map of the dataset -- known as an \textit{embedding} -- in which the distances between pairs of nearby points in the data space are preserved as far as possible.

DR techniques have found widespread use in a variety of areas of astronomy, including the analysis of low-resolution spectra.
\citet{boroson92} use PCA to explore correlations between the physical parameters of 87 low-redshift quasars observed in a small spectroscopic campaign.
\citet{richards09} employ PCA alongside diffusion maps to predict redshifts for several thousand low-redshift galaxy spectra observed by SDSS DR6.
\citet{hawkins21} use $t$SNE to reduce the dimensionality of a set of low-resolution ($R\sim750$) optical spectra of stars in the Hobby-Eberly Telescope Dark Energy Experiment (HETDEX; \citealt{gebhardt21}) to estimate a star's $\Teff$, $\log g$, and [\ce{Fe}/\ce{H}], finding 416 metal-poor candidate stars.
\citet{kao24} apply UMAP to low-resolution ($R\sim50$) Gaia DR3 XP coefficients of white dwarf spectra, isolating a group of 465 objects of which 90 are known \textit{polluted} WDs, showing metal absorption features; the remaining 375 candidates are subject to an ongoing high-resolution spectroscopic campaign.
DR has however not been extensively applied to datasets of higher-resolution spectra, perhaps due to the much larger number of wavelength bins, and hence much higher dimensionality.

White dwarfs (WDs) are the final state of the $\gtrsim97$ per cent of stars with a zero-age mass of less than $9$--$12\,\mathrm{M}_\odot$ \citep{lauffer18, althaus10, althaus21}.
Without internal nuclear fusion, these stellar cinders gradually cool on well-characterised timescales, making them important tracers of the evolution and assembly of the Galaxy (e.g., \citealt{winget87, tremblay14}).
The spectrum of a white dwarf usually deviates from that of a black body due to absorbing species in its atmosphere.
With a typical mass of $\sim0.6\,\mathrm{M}_\odot$ and radius of $\sim \mathrm{R}_E$, the extreme gravitational field gives WDs a strongly stratified structure.
Metals diffuse down through WD atmospheres on timescales much shorter than their cooling age \citep{schatzman45, paquette86, koester09, wyatt14}, leaving thin atmospheres dominated by H and/or He.
Most WD spectra show absorption features due to these light elements, and a WD is classified respectively as DA or DB if spectral features of H or He are present.
As these WDs cool below particular temperatures, atomic transitions cease to be excited ($\lesssim5000\,\text{K}$ for H; $\lesssim 11000\,\text{K}$ for He) and the absorption lines gradually fade away, giving a featureless spectrum (DC).
In certain WDs, the convection zone reaches deep into the interior and dredges up carbon from the inert core, imprinting molecular \ce{C2} Swan bands in the spectrum (DQ; \citealt{fontaine84, koester82, koester20, blouin23}).
In other cases, instabilities in remnant planetary systems can cause planetesimals to be tidally disintegrated and accreted by WDs, lending metal absorption features to the spectrum \citep{bonsor11, frewen14, mustill18, maldonado20}.
These so-called \textit{polluted} WDs (DZ) provide rare insights into the bulk composition and geology of rocky exoplanets.

Following the release of the DESI EDR \citep{desiedr}, the spectra of 3673 WD candidates were visually inspected and classified by \citet{manser24}, spectroscopically confirming 2706 WDs, of which 1400 had not been previously observed spectroscopically.
The higher resolution of DESI ($R\sim2500$--$5000$) enables identification of spectral features not visible in lower-resolution spectra: of the 152 polluted WDs identified, 121 were newly discovered or previously not classified as polluted.
However, visual classification requires significant amounts of expert time, and substantial increases in target numbers in the future ($\sim70\,000$ WD candidates from the DESI first data release; \citealt{cooper23}) all but necessitates the use of automated methods.

One class of methods for automatically classifying and analysing WD spectra that has achieved success is supervised machine learning (for example, \citealt{yang20, tan23, garciazamora23, vincent23, vincent24}).
However, a drawback of supervised machine learning techniques is the reliance on the training set used to train models.
The `imbalanced learning problem' describes the well-documented difficulties of training supervised classifiers on datasets with large class imbalances (e.g., \citealt{he09, johnson19}).
Such class imbalances are inevitable in WD datasets as some spectral types are naturally rarer than others.
Additionally, incorrect labels in the training set can severely limit the accuracy of the resulting model (e.g., \citealt{frenay14}).
Unsupervised procedures such as dimensionality reduction do not rely on an external training set; they merely reveal structures present within the dataset itself.

In this work, we apply dimensionality reduction -- specifically, $t$SNE -- to intermediate-resolution white dwarf spectra from the DESI EDR, outlining the method's success in characterising WD spectra in an unsupervised way.
We emphasise that these methods could equally be applied to sets of main-sequence stars, quasars, galaxies, or any other set of sources targeted by the aforementioned spectroscopic surveys.
The paper proceeds as follows.
In Section~\ref{sec:methods}, we describe the dataset and dimensionality reduction in more detail, as well as important preprocessing considerations.
In Section~\ref{sec:results}, we demonstrate this method's ability to identify clusters and sequences in the DESI EDR WD catalogue.
We also outline a way to incorporate prior knowledge about the locations of distinctive spectral lines, and a method of using DR in a supervised manner to classify spectra external to the catalogue.
Section~\ref{sec:discussion} discusses the application of this method as an aid in the classification of large spectroscopic WD datasets.
Section~\ref{sec:conclusions} summarises our work.

@@

The DESI EDR contains $N=3673$ WD candidates from the catalogue compiled by \citet{gentilefusillo19} where the exposures include a median signal-to-noise ratio $>0.5$ in at least one of DESI's three spectral arms \citep{manser24}.
These publicly available\footnote{
    \url{https://data.desi.lbl.gov/public/index.html}
} exposures were obtained and restacked.
The spectra span $3600$--$9824\,\text{\AA}$, with a wavelength spacing of $\Delta \lambda = 0.8\,\text{\AA}$, giving $D=7781$ flux bins for each spectrum\footnote{
    Note that the full-width-half-maximum line width of the DESI spectrographs is $\approx1.8\,\text{\AA}$ \citep{abareshi22}; these data are therefore slightly oversampled.
}.

@@

A spectrum can be represented by a high-dimensional vector, with each component of the vector corresponding to the flux in a given wavelength bin.
Spectra produced by the DESI pipeline have 7781 bins, so they can be mapped one-to-one to 7781-dimensional vectors.
These vectors are not uniformly-distributed in $\mathbb{R}^{7781}$;
spectra with shared spectral features are clustered together, as their vector components corresponding to wavelength bins near shared spectral features will take similar values.
For example, the 3703rd component corresponds to a wavelength of $6562.4\,\text{\AA}$, very close to the H$\,\alpha$ line.
The (normalised) spectra of DA WDs will have lower values in their 3703rd component than spectra without any H absorption features.
DAs will hence -- in this component, and by extension other components -- be close together.
Additionally, in the absence of artefacts or large noise, adjacent components of a given spectrum are likely to have similar values.
The components of spectra are thus highly intercorrelated.
This inherent structure in the data means that each point in the dataset lies on or close to a submanifold of the data space with much lower dimensionality.
DR techniques seek to find a low- (usually two-) dimensional representation of the structure of the high-dimensional dataset.

A popular non-linear dimensionality reduction technique is $t$-distributed Stochastic Neighbour Embedding ($t$SNE; \citealt{tsne}).
Briefly, $t$SNE attempts to map a set of high-dimensional vectors $\xb_1, \xb_2, \dots, \xb_N \in \mathbb{R}^D$ into a set of two-dimensional embedding vectors $\yb_1, \yb_2, \dots, \yb_N \in \mathbb{R}^2$, in such a way as to preserve the `similarity' between each pair of data points.
The details of how this is achieved are given in Appendix~\ref{sec:app} (see also \citealt{tsne}).
Many other DR techniques, such as Uniform Manifold Approximation and Projection (UMAP; \citealt{umap}) and LargeVis \citep{largevis} can be shown to belong to the same family of procedures, but with differing definitions of similarity (see Appendix C of \citealt{umap}).

@@

Before applying DR, it is of paramount importance to preprocess the spectra.
The spectra show a range of absolute scales: the median fluxes span three orders of magnitude.
A na\:{i}ve application of DR to the raw data would lead to two identical WDs at different distances (and hence different brightnesses) to be embedded far away from each other.
As we are interested in the intrinsic properties of the WDs in this sample, rather than their distance from Earth, the spectra must be normalised before DR is applied.
Additionally, some spectra show large spikes, due to cosmic rays or detector artefacts (see Fig.~\ref{fig:preprocessing}).
Interpreted as vectors, this corresponds to some components of the vector being very large, and as a result the vector would be very far away in data space from where it would be without the artefacts.

The preprocessing we apply to the data is illustrated in Fig.~\ref{fig:preprocessing}.
Artefacts usually correspond to pixels with very low signal-to-noise: pixels with $\mathrm{S}/\mathrm{N}<0.2$ are linearly interpolated.
Fig.~\ref{fig:preprocessing} shows that major, though not all, artefacts are removed by this step.
Overzealous removal of artefacts could remove genuine spectral features, such as the sharp emission lines of cataclysmic variables (see Section~\ref{sec:CVs}): we found that the interpolation described improved the effectiveness of the subsequent dimensionality reduction as compared to, say, median boxcar smoothing.
Following the removal of artefacts, the spectra are normalised to zero mean and unit variance.

\begin{figure*}
\centering
\includegraphics[width=0.97\textwidth]{figures/fig1_preprocessing.pdf}
\caption{
    Preprocessing stages, as illustrated on a cherry-picked DESI EDR spectrum.
    The upper panel shows the raw spectrum, with several artefacts.
    The second panel shows that the signal-to-noise ratio is very low near many of these artefacts; where it falls below 0.2, the pixels are interpolated.
    The spectra are then rescaled to zero mean and unit variance, as shown in the lower panel.
    Major artefacts have been removed, though some smaller artefacts remain.
}
\label{fig:preprocessing}
\end{figure*}

In its simplest form, dimensionality reduction treats all components of the vectors equally.
However, some wavelengths are of more significance than others, namely those of atomic absorption and emission features.
To `focus' the technique on particular spectral lines, we crop the spectrum to an appropriate window before applying DR.
This additional preprocessing step is explored in Section~\ref{sec:zoom}.

@@

Dimensionality reduction was applied to the spectra of the WDs in the DESI EDR WD catalogue, using $t$SNE.
The dimensionality reduction takes an average of $5.5\,\text{s}$ on an Intel $3.60\,\text{GHz}$ CPU.
The resulting embedding is shown in Fig.~\ref{fig:full_spectra}(a).
Each point in this embedding corresponds to an individual spectrum, and several clusters and sequences are apparent.
There is a long V-shaped sequence stretching from the left, to the bottom, to the upper right of the embedding; it turns off into a cluster towards the centre right.
A shorter secondary sequence extends from the top of the embedding towards the centre.
Several strings and clumps are found in between and around these two sequences.

The nature of these structures is elucidated by Fig.~\ref{fig:full_spectra}(b) and (c), showing respectively the primary spectral class (according to \citealt{manser24}) and effective temperature (acc. H-atmosphere model fitting of \citealt{gentilefusillo19}).
Several interesting features include:
\begin{itemize}
\item
The V-shaped sequence consists of largely DA white dwarfs;
\item
The secondary sequence in the upper part of the frame contains DBs, DZs, DQs, and DCs;
\item
The WDs in the DA sequence have been sorted approximately according to effective temperature, with hotter WDs towards the left of the sequence\footnote{
    This trend is also apparent in the secondary sequence, but because many of these are DBs and DQs, and the temperatures shown here are fitted assuming H-atmosphere models, the trend is less meaningful.
    However, the trend is nonetheless borne out in the fitted temperatures of He-atmosphere models.
};
\item
DZs are found in various places\footnote{
    \citet{kao24} also find that DZs with different characteristics (primarily $\Teff$) are scattered by dimensionality reduction into multiple regions.
}: (i) an island near to the secondary sequence, (ii) towards the cool end of the DA sequence, (iii) amongst the secondary sequence;
\item
Extragalactic sources (`$\times$' markers) are mostly located in strings and small clumps around the secondary sequence;
\item 
Subdwarfs (`$+$' markers) are largely found towards the hot end of the DA sequence;
\item
The majority of main-sequence stars (black star icons) are found towards the centre right, though their cluster merges somewhat with the cool end of the DA sequence.
A few are also scattered elsewhere.
\end{itemize}

>>>>

% Abstract of the paper
\begin{abstract}
As a new generation of large-sky spectroscopic surveys comes online, the enormous data volume poses unprecedented challenges in classifying spectra.
Although supervised classifiers exist, their complete reliance on training data can lead to overconfident predictions and bias against rarer classes.
We outline the use of dimensionality reduction to generate a 2D map of the structure of an intermediate-resolution spectroscopic dataset.
We demonstrate that this unsupervised technique efficiently separates white dwarfs of different spectral classes in the Dark Energy Spectroscopic Instrument's Early Data Release (DESI EDR), identifying spectral features that had been missed even by visual classification.
By zooming in to important wavelength regions, we show how focusing the method on particular spectral features enables the identification of helium lines with 90 per cent accuracy, and cataclysmic variables with 100 per cent accuracy.
We also demonstrate a way of using dimensionality reduction in a supervised manner, outlining a procedure to classify any white dwarf spectrum in comparison with those in the DESI EDR.
With upcoming surveys promising tens of millions of spectra, this work highlights the potential for unsupervised techniques as efficient means of classification and dataset visualisation.
\end{abstract}

@@

A suite of upcoming large-sky spectroscopic surveys promises to enormously increase the number of sources with intermediate-resolution spectroscopy, enabling advances across all areas of astronomy.
These surveys include:
\begin{itemize}
\item
\textbf{4MOST}
(4-metre Multi-Object Spectroscopic Telescope; \citealt{4most}): nine surveys which aim to obtain accurate radial velocities and chemical abundances for millions of Galactic stars, probe the history of supermassive black hole accretion through a spectroscopic sample of about 1 million active galactic nuclei, and constrain cosmological parameters by obtaining redshifts for 10-20 million emission line galaxies;
\item
\textbf{DESI}
(Dark Energy Spectroscopic Instrument; \citealt{desii, desiii}), which plans to study the growth of structure in the Universe through observations of baryon acoustic oscillations and redshift-space distortions, and trace the dark matter distribution by obtaining redshifts for 30 million galaxies and quasars at $1.0\lesssim z \lesssim 3.5$.
When observations of these faint targets are hindered by moonlight or poor observing conditions over Iolkam Du'ag / Kitt Peak, the targeting switches to surveys of Milky Way stars and bright galaxies;
\item
\textbf{SDSS-V}
(Sloan Digital Sky Survey V; \citealt{sdssv}):
an all-sky survey from two $2.5\,\text{m}$ telescopes (one in each hemisphere), this aims to record multi-epoch spectra for over six million targets, including five million Galactic stars;
\item
\textbf{WEAVE}
(William Herschel Telescope Enhanced Area Velocity Explorer; \citealt{weave}): a northern-hemisphere complement to 4MOST, WEAVE will record accurate spectroscopic velocities of Galactic disk and halo stars, IFU H~\textsc{i} maps of $\sim10^4$ low-redshift galaxies, and wide-area views of galaxy clusters.
\end{itemize}
All four of these surveys have achieved first light, with SDSS-V's first data release (SDSS DR19; \citealt{sdssdr19}) and DESI's early data release (DESI EDR; \citealt{desiedr}) delivered last year.
As these spectroscopic surveys begin to return enormous quantities of data, and with the exabyte-scale Legacy Survey of Space and Time (LSST; \citealt{lsst}) on the horizon, automated techniques will become absolutely necessary in aiding the extraction of scientific results from the vast wealth of data being collected.

Entries in astronomical datasets often have high dimensionality.
They may be image cutouts of hundreds of pixels, light curves with thousands of measurements, or spectra with thousands of fluxes.
Any \textit{structure} in a dataset -- e.g., clusters or sequences -- can be investigated quantitatively by representing each entry as a vector in a very high-dimensional space: a dataset of $N$ spectra with $D$ flux bins can be thought of as a set of $N$ points in $D$-dimensional data space.
Visualisation of datasets in more than two or three dimensions is challenging, but whereas a $D$-dimensional data point might ostensibly live in $\mathbb{R}^D$, it is very likely that every point in the dataset is located on or near a much lower-dimensional submanifold within this space.
In a spectrum, the flux in a given bin will usually be similar to the flux in adjacent bins.
Similarly, two different sources of the same astrophysical type (e.g.\ two white dwarfs of the same spectral classification) will have similar spectra, and will thus be nearby in the data space.
This massively reduces the effective dimensionality of the dataset, compared to the dimensionality of the raw data points themselves.
The observation that many datasets exist near low-dimensional submanifolds of the data space has motivated the development of several \textit{dimensionality reduction} techniques, including Principal Component Analysis (PCA; \citealt{pca}), diffusion maps \citep{coifman06, lafon06}, Locally Linear Embedding (LLE; \citealt{roweis00}), $t$-distributed Stochastic Neighbour Embedding ($t$SNE; \citealt{tsne}), and Uniform Manifold Approximation and Projection (UMAP; \citealt{umap}).
These methods calculate a two-dimensional map of the dataset -- known as an \textit{embedding} -- in which the distances between pairs of nearby points in the data space are preserved as far as possible.

Dimensionality reduction techniques have found widespread use in a variety of areas of astronomy, including the analysis of low-resolution spectra.
\citet{boroson92} use PCA to explore correlations between the physical parameters of 87 low-redshift quasars observed in a small spectroscopic campaign.
\citet{richards09} employ PCA alongside diffusion maps to predict redshifts for several thousand low-redshift galaxy spectra observed by SDSS DR6.
\citet{hawkins21} use $t$SNE to reduce the dimensionality of a set of low-resolution ($R\equiv \lambda / \Delta \lambda \sim750$) optical spectra of stars in the Hobby-Eberly Telescope Dark Energy Experiment (HETDEX; \citealt{gebhardt21}) to estimate a star's effective temperature ($\Teff$), gravitational field strength ($\log g$), and metallicity, finding 416 metal-poor candidate stars.
\citet{kao24} apply UMAP to low-resolution ($R\sim50$) Gaia DR3 XP coefficients of white dwarf spectra, isolating a group of 465 objects of which 90 are known \textit{polluted} white dwarfs, showing metal absorption features; the remaining 375 candidates are subject to an ongoing high-resolution spectroscopic campaign.
Dimensionality reduction has however not been extensively applied to datasets of higher-resolution spectra, perhaps due to the much higher dimensionality of spectra with many more wavelength bins.

White dwarfs (WDs) are the final state of the $\gtrsim97$ per cent of stars with a zero-age mass of less than $9$--$12\,\mathrm{M}_\odot$ \citep{lauffer18, althaus10, althaus21}.
Without internal nuclear fusion, these stellar cinders gradually cool on well-characterised timescales, making them important tracers of the evolution and assembly of the Galaxy (e.g., \citealt{winget87, tremblay14}).
The spectrum of a white dwarf usually deviates from that of a black body due to absorbing species in its atmosphere.
With a typical mass of $\sim0.6\,\mathrm{M}_\odot$ and radius of $\sim \mathrm{R}_{\earth}$, the extreme gravitational field gives WDs a strongly stratified structure.
Metals diffuse down through WD atmospheres on timescales much shorter than their cooling age \citep{schatzman45, paquette86, koester09, wyatt14}, leaving thin photospheres dominated by H and/or He.
Most WD spectra show absorption features due to these light elements, and a WD is classified respectively as DA or DB if spectral features of H or He are present.
As these WDs cool below particular temperatures, atomic transitions cease to be excited ($\lesssim5000\,\text{K}$ for H; $\lesssim 11000\,\text{K}$ for He) and the absorption lines gradually fade away, giving a featureless spectrum (DC).
In certain WDs, the convection zone reaches deep into the interior and dredges up carbon from the inert core, imprinting molecular \ce{C2} Swan bands in the spectrum (DQ; \citealt{fontaine84, koester82, koester20, blouin23}).
In other cases, instabilities in remnant planetary systems can cause planetesimals to be tidally disintegrated and accreted by WDs, lending metal absorption features to the spectrum \citep{bonsor11, frewen14, mustill18, maldonado20}.
These so-called polluted WDs provide unique insights into the bulk composition and geology of rocky exoplanets.
In the absence of other features, such metal-line WDs are classified DZ; if e.g.\ H features and metal features are both present, the WD is classified as DAZ or DZA depending on which features are strongest.

Following the release of the DESI EDR \citep{desiedr}, the spectra of 3673 WD candidates were visually inspected and classified by \citet{manser24}, spectroscopically confirming 2706 WDs, of which 1400 had not been previously observed spectroscopically.
The higher resolution of DESI ($R\sim2500$--$5000$) enables identification of spectral features not visible in lower-resolution spectra: of the 152 polluted WDs identified, 121 were newly discovered or previously not classified as polluted.
However, visual classification requires significant amounts of expert time, and substantial increases in target numbers in the future ($\sim70\,000$ WD candidates from the full DESI survey \citealt{cooper23}) all but necessitates the use of automated methods.

One class of methods for automatically classifying and analysing WD spectra that has achieved success is supervised machine learning (for example, \citealt{yang20, tan23, garciazamora23, vincent23, vincent24}).
However, a drawback of supervised machine learning techniques is the reliance on the training set used to train models.
The `imbalanced learning problem' describes the well-documented difficulties of training supervised classifiers on datasets with large class imbalances (e.g., \citealt{he09, johnson19}).
Such class imbalances are inevitable in WD datasets as some spectral types are naturally rarer than others.
For example, in the DESI EDR, 1958 WDs were classified DA; only one was classified DO \citep{manser24}.
Additionally, incorrect labels in the training set can severely limit the accuracy of the resulting model (e.g., \citealt{frenay14}).
Unsupervised procedures such as dimensionality reduction do not rely on an external training set; they merely reveal structures present within the dataset itself.

In this work, we apply dimensionality reduction -- specifically, $t$SNE -- to intermediate-resolution white dwarf spectra from the DESI EDR, outlining the method's success in characterising WD spectra in an unsupervised way.
We emphasise that these methods could equally be applied to sets of main-sequence stars, quasars, galaxies, or any other set of sources targeted by the aforementioned spectroscopic surveys.
The paper proceeds as follows.
In Section~\ref{sec:methods}, we describe the dataset and dimensionality reduction in more detail, as well as important preprocessing considerations.
In Section~\ref{sec:results}, we demonstrate this method's ability to identify clusters and sequences in the DESI EDR WD catalogue.
We also outline a way to incorporate prior knowledge about the locations of distinctive spectral lines, and a method of using dimensionality reduction in a supervised manner to classify spectra external to the catalogue.
Section~\ref{sec:discussion} discusses the application of this method as an aid in the classification of large spectroscopic WD datasets.
Section~\ref{sec:conclusions} summarises our work.

We will refer to WDs with a particular primary classification as, e.g.\ `DA*', to include of DA, DAB, DAZ, etc.
Additionally, we use the broader classification `D*A' to include WDs with primary \textit{or} secondary classification, inclusive of DA, DAB, DZA, etc.

@@

The DESI EDR contains $N=3673$ WD candidates from the catalogue compiled by \citet{gentilefusillo19} where the exposures include a median signal-to-noise ratio $>0.5$ in at least one of DESI's three spectral arms \citep{manser24}.
These publicly available\footnote{
    \url{https://data.desi.lbl.gov/public/index.html}
} exposures were obtained and stacked by standard inverse variance weighting.
The spectra span $3600$--$9824\,\text{\AA}$, with a wavelength spacing of $\Delta \lambda = 0.8\,\text{\AA}$, giving $D=7781$ flux bins for each spectrum.
All spectra within DESI have an identical wavelength axis.

@@

A spectrum can be represented by a high-dimensional vector, with each component of the vector corresponding to the flux in a given wavelength bin.
Spectra produced by the DESI pipeline have 7781 bins, so they can be mapped one-to-one to 7781-dimensional vectors.
These vectors are not uniformly-distributed in $\mathbb{R}^{7781}$;
spectra with shared spectral features are clustered together, as their vector components corresponding to wavelength bins near shared spectral features will take similar values.
For example, the 3703rd component corresponds to a wavelength of $6562.4\,\text{\AA}$, very close to the H$\,\alpha$ line.
The (normalised) spectra of DA WDs will have lower values in their 3703rd component than spectra without any H absorption features.
DAs will hence -- in this component, and by extension other components -- be close together.
Additionally, in the absence of artefacts or large noise, adjacent components of a given spectrum are likely to have similar values.
The components of spectra are thus highly intercorrelated.
Dimensionality reduction techniques exploit these redundancies in datasets, seeking a low- (usually two-) dimensional representation of a dataset's structure.

A popular non-linear dimensionality reduction technique is $t$-distributed Stochastic Neighbour Embedding ($t$SNE; \citealt{tsne}).
Briefly, $t$SNE attempts to map a set of high-dimensional vectors $\xb_1, \xb_2, \dots, \xb_N \in \mathbb{R}^D$ into a set of two-dimensional embedding vectors $\yb_1, \yb_2, \dots, \yb_N \in \mathbb{R}^2$, in such a way as to preserve the `similarity' between each pair of data points.
The details of how this is achieved are given in Appendix~\ref{sec:app} (see also \citealt{tsne}).
Many other dimensionality reduction techniques, such as Uniform Manifold Approximation and Projection (UMAP; \citealt{umap}) and LargeVis \citep{largevis} can be shown to belong to the same family of procedures, but with differing definitions of similarity (see Appendix C of \citealt{umap}).

@@

Before applying dimensionality reduction, it is of paramount importance to preprocess the spectra.
The spectra show a range of absolute scales: the median fluxes span three orders of magnitude.
A na\"{i}ve application of dimensionality reduction to the raw data would lead to two identical WDs at different distances (and hence different brightnesses) to be embedded far away from each other.
As we are interested in the intrinsic properties of the WDs in this sample, rather than their distance from Earth, the spectra must be normalised before dimensionality reduction is applied.
Additionally, some spectra show large spikes, due to cosmic rays, detector artefacts, or imperfect sky subtraction (see Fig.~\ref{fig:preprocessing}).
Interpreted as vectors, this corresponds to some components of the vector being very large, and as a result the vector would be very far away in data space from where it would be without the artefacts.

The preprocessing we apply to the data is illustrated in Fig.~\ref{fig:preprocessing}.
Artefacts usually correspond to pixels with very low signal-to-noise: pixels with $\mathrm{S}/\mathrm{N}<0.2$ are linearly interpolated.
Fig.~\ref{fig:preprocessing} shows that major artefacts are removed by this step, though some sky lines and telluric features remain.
Overzealous removal of artefacts could remove genuine spectral features, particularly weaker ones: we found that the interpolation described improved the effectiveness of the subsequent dimensionality reduction as compared to, say, median boxcar smoothing.
Following the removal of artefacts, the spectra are normalised to zero mean and unit variance.

\begin{figure*}
\centering
\includegraphics[width=0.97\textwidth]{figures/fig1_preprocessing.pdf}
\caption{
    Preprocessing stages, as illustrated on the cherry-picked DESI EDR spectrum of WDJ170114.72+760207.16.
    The upper panel shows the raw spectrum, with several artefacts.
    The second panel shows that the signal-to-noise ratio is very low near many of these artefacts; where it falls below 0.2, the pixels are interpolated.
    The spectra are then rescaled to zero mean and unit variance, as shown in the lower panel.
    Major artefacts have been removed, though certain imperfect sky subtraction and telluric features remain, particularly above $7500\,\text{\AA}$.
}
\label{fig:preprocessing}
\end{figure*}

In its simplest form, dimensionality reduction treats all components of the vectors equally.
However, some wavelengths are of more significance than others, namely those of atomic absorption and emission features.
To `focus' the technique on particular spectral lines, we crop the spectrum to an appropriate window before applying dimensionality reduction.
This additional preprocessing step is explored in Section~\ref{sec:zoom}.

@@

Dimensionality reduction was applied to the spectra of the WDs in the DESI EDR WD catalogue, using $t$SNE.
The dimensionality reduction takes an average of $5.5\,\text{s}$ for the entire dataset.
The resulting embedding is shown in Fig.~\ref{fig:full_spectra}(a).
Each point in this embedding corresponds to an individual spectrum, and several clusters and sequences are apparent.
There is a long V-shaped sequence stretching from the left, to the bottom, to the upper right of the embedding; it turns off into a cluster towards the centre right.
A shorter secondary sequence extends from the top of the embedding towards the centre.
Several strings and clumps are found in between and around these two sequences.

The nature of these structures is elucidated by Fig.~\ref{fig:full_spectra}(b) and (c), showing respectively the primary spectral class (according to \citealt{manser24}) and effective temperature (acc. H-atmosphere model fitting of \citealt{gentilefusillo19}).
Several interesting features include:
\begin{itemize}
\item
The V-shaped sequence consists of largely DA* white dwarfs;
\item
The secondary sequence in the upper part of the frame contains DB*s, DZ*s, DQs, and DCs;
\item
The WDs in the DA sequence have been sorted approximately according to effective temperature, with hotter WDs towards the left of the sequence;
\item
DZ*s are found in various places\footnote{
    \citet{kao24} also find that DZs with different characteristics (primarily $\Teff$) are scattered by dimensionality reduction into multiple regions.
}: (i) an island near to the secondary sequence, (ii) towards the cool end of the DA sequence, (iii) amongst the secondary sequence;
\item
Extragalactic sources (`$\times$' markers) are mostly located in strings and small clumps around the secondary sequence;
\item 
Subdwarfs (`$+$' markers) are largely found towards the hot end of the DA sequence;
\item
The majority of main-sequence stars (black star icons) are found towards the centre right, though their cluster merges somewhat with the cool end of the DA sequence (cf.\ fig.~1 of \citealt{eisenstein06}).
A few are also scattered elsewhere.
\end{itemize}

\begin{figure*}
    \centering
\includegraphics[width=0.5\textwidth]{figures/fig2a_embedding.pdf}
\includegraphics[width=0.99\textwidth]{figures/fig2b_embedding.pdf}
\caption{
    (a) $t$SNE embedding of DESI EDR spectra.
    The dimension-reduced embedding aims to reflect the pairwise distances between the high-dimensional spectra in a two-dimensional space; as such the axes are arbitrary.
    (b) The embedding is colour-coded according to the visual spectral classification of \citet{manser24}.
    For WDs, the colour corresponds to primary spectral type: DA*s (inc.\ DAZs, etc.) in red; DB*s in blue, and so on.
    Other sources not corresponding to individual WDs have different symbols (see key).
    The main feature of the embedding is the sequence of DA*s (red), though several other clusters are clear (see text).
    (c) The embedding is colour-coded by effective temperature, according to the hydrogen-atmosphere WD model which best fits the sources' Gaia photometry \citep{gentilefusillo19}.
    The DA sequence extends from hotter WDs on the left around to cooler WDs at the top right.
    Note that objects with a low probability of being a WD ($P_\mathrm{WD} < 0.75$; see \citealt{gentilefusillo15}) are not assigned a temperature \citep{gentilefusillo19}.
}
\label{fig:full_spectra}
\end{figure*}

Although the spectral classes are known here, owing to the visual inspection campaign of \citet{manser24}, dimensionality reduction would nonetheless provide a valuable analysis if the classes were \textit{a priori} unknown, as will be the case with upcoming spectroscopic surveys.
For example, it is well-known that approximately 80 per cent of WDs are DAs.
As such, when applying dimensionality reduction to a WD dataset, one would implicitly identify the largest cluster as corresponding to mostly DAs.
